Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference

### 2017-06-29 workflow/workflow.Rmd
Updates to text.

---
title: "Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, UC Berkeley
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
  - name: Kelly N. Street
    affilitation: Graduate Group in Biostatistics, UC Berkeley
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, UC Berkeley
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics, UC Berkeley, and Department of Statistics, UC Berkeley
abstract: Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized.
keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, clustering, lineage inference, statistical models, workflow, R, Bioconductor
bibliography: ref.bib
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
---

@@

@@

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis within and between the lineages. Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.

@@

[SD: Fig. 1: For DE, might want to deemphasize edgeR, as it is not our contribution, by mentioning other packages.]

@@

Using the Bioconductor R package `scone` [SD: scone not yet released on Bioconductor], we remove low-quality cells that did not pass our quality control filter, that is ... See Figure \@ref(fig:scone). Davide, would you like to describe your choices for scone here?

@@

The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step. We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`. Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., `k`-means, with a range of values for `k`), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005]. A consensus clustering is obtained based on the co-clustering matrix and non-differential clusters are then merged by creating a hierarchy of clusters, working up the tree, testing for differential expression between sister nodes, and collapsing nodes with insufficient DE. As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.

@@

To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`. We found that the slingshot results were robust to the number of dimensions `k` for the MDS (we tried `k=` [SD: Add values of `k` that were considered.]). Here, we use the unsupervised version of slingshot, where we only provide the identity of the start cluster but not of the end clusters. This implementation recovers the lineages found in the published work (Figures \@ref(fig:curves) and \@ref(fig:tree)).

@@

# Software availability
This section will be generated by the Editorial Office before publication. Authors are asked to provide some initial information to assist the Editorial Office, as detailed below.

@@

The four packages used in the workflow (scone, zinbwave, clusterExperiment, and slingshot) are bioconductor packages and are available on github at respectively https://github.com/YosefLab/scone, https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot. The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow under license XXX.

@@

FP, DR, EP, and KNS performed the data analaysis. DR, SD, and EP supervised the research. All authors contributed to the writing of the manuscript.

@@

DR, KNS, EP, and SD were supported by the National Institutes of Health BRAIN Initiative (grant U01 MH105979, PI: John Ngai). KNS was supported by a training grant from the National Human Genome Research Institute (T32000047).

# Acknowledgments
This section should acknowledge anyone who contributed to the research or the
article but who does not qualify as an author based on the criteria provided earlier (e.g. someone or an organization that provided writing assistance). Please state how they contributed; authors should obtain permission to acknowledge from all those mentioned in the Acknowledgments section.

>>>>

---
title: "Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, UC Berkeley
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
  - name: Kelly N. Street
    affilitation: Graduate Group in Biostatistics, UC Berkeley
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, UC Berkeley
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics, UC Berkeley, and Department of Statistics, UC Berkeley

abstract: Novel single-cell transcriptome sequencing assays allow researchers to measure gene expression levels at the resolution of single cells and offer the unprecendented opportunity to investigate at the molecular level fundamental biological questions such as stem cell differentiation or the discovery and characterization of rare cell types. However, such assays raise challenging statistical and computational questions and require the development of novel methodology and software. Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages.

keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, clustering, lineage inference, statistical models, workflow, R, Bioconductor

bibliography: ref.bib

vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
---

@@

[SD: Fix double affiliation for SD so that there are 2 numbers.]

@@

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis alonglineages. Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.

@@

[SD: Fig. 1: For DE, remove mention to edgeR.]

@@

Using the Bioconductor R package `scone`, we remove low-quality cells that did not pass our quality control filter, that is ... See Figure \@ref(fig:scone). Davide, would you like to describe your choices for scone here?

@@

The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step. We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`. Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005]. A consensus clustering is obtained based on the co-clustering matrix and non-differential clusters are then merged by creating a hierarchy of clusters, working up the tree, testing for differential expression between sister nodes, and collapsing nodes with insufficient DE. As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.

@@

To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`. We found that the slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_= [SD: Add values of _k_ that were considered.]). Here, we use the unsupervised version of slingshot, where we only provide the identity of the start cluster but not of the end clusters. This implementation recovers the lineages found in the published work (Figures \@ref(fig:curves) and \@ref(fig:tree)).

@@

# Software availability

This section will be generated by the Editorial Office before publication. Authors are asked to provide some initial information to assist the Editorial Office, as detailed below.

@@

The four packages used in the workflow (scone, zinbwave, clusterExperiment, and slingshot) are Bioconductor packages and are available on github at respectively https://github.com/YosefLab/scone, https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot. The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow under license XXX.

@@

FP, DR, KNS, and EP performed the data analaysis and wrote the code portions of the workflow. FP and SD wrote the text portion of the workflow, with contributions from the other three authors. DR, EP, and SD supervised the research.

@@

DR, KNS, EP, and SD were supported by the National Institutes of Health BRAIN Initiative (U01 MH105979, PI: John Ngai). KNS was supported by a training grant from the National Human Genome Research Institute (T32000047).

# Acknowledgments

The authors are grateful to John Ngai (Department of Molecular and Cell Biology, UC Berkeley) and his group members Russell B. Fletcher and Diya Das for motivating the research presented in this workflow and for valuable feedback on applications to biological data.
